\documentclass[a4paper,12pt]{article}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

\hypersetup{
  pdftitle={},
  pdfauthor={Kimmo Puputti},
  pdfsubject={},
  pdfkeywords={},
  pdfcreator={pdflatex},
  pdfproducer={pdflatex},
  colorlinks=true,
  linkcolor=blue,
  citecolor=green,
  filecolor=magenta,
  urlcolor=blue
}


\title{}
\author{\textbf{Kimmo Puputti}\\kimmo.puputti@futurice.com\\\url{http://kpuputti.fi/}}

\begin{document}

\maketitle
\clearpage

% Introduction
\section{Introduction}

\begin{itemize}
\item The Web is integrated to our daily lives
\item The Web is accessed (for consuming and producing information)
  from everywhere with a ton of different devices
\end{itemize}

\noindent Principles of the Web:

\begin{itemize}
\item The Web is ubiquitous
\item The Web is universal
\item The Web should be open
\item The Web should be accessible for anyone, anywhere, anytime
\item Resources on the Web should be linkable by anyone from anywhere
\end{itemize}

\noindent Native apps are popular, but being cross-platform is
essential. Smart phones and tablets are everywhere.

% Research Question
\section{Research Question}

\begin{itemize}
\item What HTML5?
\item Why HTML5?
\item How HTML5?
  \begin{itemize}
  \item How to handle flaky network connections?
  \item How to make apps blazing fast?
  \item How to make the apps responsive?
  \item How to attack different screen sizes?
  \item How to achieve cross-platform support?
  \end{itemize}
\item Promises and hype are big, but what are the realities and
  real-world performance?
\end{itemize}

% Methods
\section{Methods}

We created the Qt Developer Days 2011 conference mobile website for
schedule management and session information and rating. The website
has offline support, stores persistent user-specific data in the
client side HTML5 localStorage and scales to different screen sizes.

% Results
\section{Results}

We created a cross-platform website for mobile phones and tablets
running on several different platforms. Design compromises were made
to achieve cross-platform support.

% Discussion
\section{Discussion}



\end{document}
